Data Specialization for a Staged Lambda Calculus

### 2014-11-11 paper/example.tex
cleaning up latex warnings
major reorganization of section 2 (in progress) - moving lots of text around!

\subsection{Comments, suggestions, and ``todo''}
I rewrote this section.  There is more work to be done.

\begin{itemize}
\item 
I find it hard to distinguish between blue and black on printed paper.

\item
$\pause$ should be called ``promote'' or something like that. I
  believe this is what Taha calls cross stage persistence.  THat work should
  be cited.

\item
I don't like fuzzy words like hygenic that dont mean anything.
  What is the point of using them?  A paper ought to be written for
  the reader not as a note to self.

\item the splitting example should shed more light to the techniques
  and try to give more insight.
\end{itemize}

\begin{figure*}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) 
  : (int*list*list) =
  case unroll l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) =>
      let (s,left,right) = partition (p,t) in
      if h<p 
      then (s+1,Cons(h,left),right)
      else (s,left,Cons(h,right))

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\label{fig:quickselect}
\label{fig:qs-unstaged}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
  case l of
    Empty => next {`2`0`1`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        `2`let n = hold{`1`n`2`} in
          case compare prev{`1`k`2`} n of
            LT => prev {`1`qsStaged (left k)`2`}
          | EQ => hold {`1`h`2`}
          | GT => prev {`1`qsStaged (right, 
                             next{prev{k}-n-1)}`2`}`1`
      }`
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\label{fig:qsstaged}
\label{fig:qs-staged}
\end{subfigure}
\caption{Quickselect: traditional and staged.}
\end{figure*}

Staging is a technique for using language constructs and types to
explicitly staged a computation according to the frequency or the
arrival time of data.  As an example, suppose that we wish to issue
order statistics queries on a collection of items represented as a
list \texttt{l}.  For this tasks, we can use the expected linear time,
quickselect algorithm~\cite{quickselect}, which takes a
list~\texttt{l} and a rank~\texttt{k} and returns the element with
rank~\texttt{k} in the list.  To this end, the algorithm, whose code
is shown in ~\ref{fig:qs-unstaged} partitions the list into two by
using the first element as a pivot\footnote{We assume that the list
  is prepermuted to guarantee the expected linear time behavior.} and
recurs down to one of the ``halves'' to find the desired element.  The
``half'' to be considered recursively next is determined by the
relationship of~\texttt{k} to the size of the first half~\texttt{n},
which is returned by the function~\texttt{partition}.

Continuing on our example, consider an application where we perform
many order statistics queries using the same collection~\texttt{l}, but
 with $m$ different ranks~\texttt{k}:
%
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@.
\end{lstlisting}
%
Since \texttt{quickselect1} requires only on a first stage
argument~\texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on~\texttt{l},
e.g., the code consisting of $m$ calls To take advantage of this, we
can {\em stage} quickselect so that the parts of the computation that
depend on the collection~\texttt{l} is performed once, and the desired
result is computed by using~\texttt{k} and the results from this prior
stage.

Prior work proposed staging for exactly these kinds of applications
and proposed techniques for writing staged
programs~\cite{staging}. Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \ref{fig:qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage.

Having determined the type signature,
the programmer can then insert $\next$ and $\prev$ constructs into the
code. The construct $\next$ occurs in a stage \bbone\ context and
indicates that the contents of its block are stage \bbtwo, whereas
$\prev$ occurs in a stage \bbtwo\ context and indicates that the
contents of its block is stage \bbone.  (We adopt the convention that
the top-level context is stage \bbone.)  The output type of a $\next$
block is the $\fut$'d version of the type of its stage
\bbtwo\ contents.  For example, \verb|next{0}| from above has the type
$\fut\rmint$.  Correspondingly, $\prev$ requires that its stage
\bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage
\bbtwo, since $k$ is bound to a $\fut\rmint$ at stage
\bbone. Informally speaking, the
$\next$ construct enables operating on second stage computations in
the first stage essentially by treating them as boxes that can be
moved around, stored in data structures, and passed to functions, and
the $\prev$ construct enables unwrapping the contents of such a box to
use in the second stage.

For convenience, the code also uses $\pause$ blocks.  This construct
has same stage signature as $\prev$, but instead of ``unwrapping"
$\fut$ types it simply promotes integers from stage~\bbone\ to
stage~\bbtwo.
\footnote{It will turn out that $\pause$ is implementable---though it
  takes some effort---from our other language features.  We instead
  provide it as a primitive to shorten examples.  Furthermore, it
  would be wise to extend $\pause$ to other base types, if we had
  them, and to products and sums thereof.  This is related to the
  notion of {\em mobility} in \cite{murphy05} and {\em stability} in
  \cite{krishnaswami13}.} We note that given the type signature of
staged quickselect, there are several ways to insert the staging
annotations.  The version we presented here maximizes the work done at
stage~\bbone.

%%%
%% There has been the extensive research into the question of how to
%% automatically add staging annotations to unstaged code.  This process
%% is known as {\em binding time analysis}, and we do not consider it
%% here.  Instead, we assume that all input programs are properly staged
%% according to some programmer's intent.

Prior work on meta programming allows us to evaluate such staged code
in a staged fashion by first providing the first-stage arguments and
generating a piece of code that is specialized for the arguments.  For
example, we can evaluate the staged quickselect by providing a
list~\texttt{l} to obtain a piece of code that then can be used with
different ranks (\texttt{k} values) in the second stage.  Such a
specialized code essentially performs all the first-stage computations
first and uses their results to improve the efficiency in the second
stage, where we issue many queries.  The type system guarantees that
all first-stage computations depend only on the first stage values and
can indeed be computed in the absence of second-stage values.

This staged evaluation technique corresponds to {\em partial
  evaluation} or {\em program specialization} and can bring out
dramatic improvements in the efficiency of programs.  It has however
one major limitation: since it requires the first stage arguments to
be available, it cannot be performed at compile time, except of course
when the first-stage values are available at compile time.  Since the
evaluation in the first stage produces a piece of code that must also
be compiled or interpreted to execute in the second stage, the
approach can lead to code blowup and tends to be relatively expensive
in practice.  Due to these reasons, the applications of partial
evaluation are limited.  For example, in our quickselect example, it
is unlikely to know the list~\texttt{l} that we will be operation on
at compile time. Partial evaluation dynamically at run time is
difficult because the first stage generates code that is linear in the
size of the list, whose generation, dynamic compilation or
interpretation can be excruciatingly slow.

Knoblock and Ruf~\cite{KR} proposed the {\em data specialization}
technique to overcome some of these limitations of program
specialization.  The idea behind this technique is to take a staged
program and split it into a stage-1 and a stage-2 program without
knowledge of any of the arguments.  Unlike program specialization,
data specialization does not require knowledge of the first-stage
arguments to produce fast code for the second stage.  Instead it yield
a first stage program that maps the first stage arguments to an
intermediate or {\em boundary data structure}, which is used by the
second stage program along with the second stage arguments to produce
the final results.  Knoblock and Ruf's observation is that the
information required by the second stage can be represented solely as
a data structure, making it unnecessary to generate and compile or
interpret code dynamically at run time.  Data specialization can
therefore be applied statically at compile time, leading to
drastically improved run-time performance without any significant
run-time overheads.

By applying data specialization to our quickselect example, we would
wish to derive two functions, \texttt{quickselect1} and
\texttt{quickselect2}, that would be semantically equivalent to
quickselect in the following sense:
\\
\texttt{quickselect l k = quickselect2 (quickselect1 l) k}.
\\
%
Since \texttt{quickselect1} requires only on a first stage
argument~\texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on~\texttt{l},
e.g.,  the code consisting of $m$ calls 
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@
\end{lstlisting}
%
can be replaced by 
%
\begin{lstlisting}
b = quickselect1 l
quickselect2 b 1
quickselect b 2
@$\ldots$@ 
quickselect b @$m$@.
\end{lstlisting}

The benefit of this ``optimization'' can be dramatic.  For example, in
principle at least, the first stage function \texttt{quickselect1}
could produce a balanced binary search tree holding the elements
of~\texttt{l} and the second stage function \texttt{quickselect2}
could use this balanced binary search tree and the given rank value to
perform a logarithmic-time search in the tree to return the desired
result.  Assuming that the list~\texttt{l} contains $n$ elements, this
optimization would change the asymptotic complexity from expected
(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the complexity
from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.  Note that the desired optimized code shows above is
intellectually more sophisticated than the code that we have started
with: the optimized code is able to create a data structure, a
balanced binary tree augmented with indexing information, and use a
binary search technique over this tree to compute the result
asymptotically more efficiently.  In fact, based our teaching
experience, we can imagine this kind of problem to be a moderately
difficult exam question in an undergraduate algorithms class, as it
not only requires understanding of data structures such as binary
search trees but also requires modifying them to augment with indexing
information to support rank-based search.

It is thus perhaps no surprise that the prior art in data
specialization is not able to perform such ``optimizations''.
Knoblock and Ruf's paper takes a very important step by formulating
the problem and proposing a solution.  But their work applies to
program fragments that don't involve sophisticated data types such as
as sums and recursive types, and don't involve recursion. For example,
their work would not apply to our quickselect example and similar ones
involving recursion and/or  recursive types.

@@

In what follows we present a relatively mechanical application of our
techniques to our running example.

\ur{I made up to here.  What follows should be shortened and made to
  the point. Some time later...}

To illustrate this, we present colorized version of \texttt{qSelect}:

\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ... `

fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => `qSelect 1`left `2`k
      | EQ => h
      | GT => `qSelect 1`right `2`(k-n-1)`
\end{lstlisting}

Here, the input $k$ and all parts of the computation that depend on it are colored blue,
whereas all of the parts of the computation that do not depend on $k$ are colored red.
The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.

\ur{I find control dependence issue very confusing.  Let's factor this
out and talk about it in the end. }

To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
first arguments to the recursive calls to \texttt{qSelect} are colored red,
despite appearing underneath a blue case statement.
This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.

Observe the that helper function \texttt{partition} is colored entirely red,
which is valid because it is called with only red arguments.  
This is an illustration of the property stated above, that 
the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.

This suggests an optimization: we can split \texttt{qSelect} into two functions, 
one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
Because of the dependence properties stated above, 
\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
whereas \texttt{qSelect1} will only depend on the input list $l$.
Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
Equationally, we can represent this relationship as simply 
\[
\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
\]
for some list \texttt{l} and index \texttt{k}.

\ur{I don't follow the explanation below, e.g., ``assuming conclusion''}

With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
Because the original function is recursive,
the first step in this process is to assume the conclusion at the call sites.

\ur{quickselect 1 and 2: how did we introduce them?}

This is performed in place:
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}
We then pull the stage \bbone\ portion of each branch of the
comparison up into a single tuple.
\ur{Why did we include h}

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}

\ur{Don't know what ``this last step'' refers to.}
This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
We continue to pull up the precomputation and include more in it:

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

This completes the splitting of the cons branch.  
The empty branch can also be split, though in a trivial way:

\ur{Not clear, ``can be split'' split.  Does or doesn't (split) should be the verb?}
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => let p = () in `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

To continue splitting the case statement, we need to generate a sum type in the precomputation
to remind the code what branch its in.

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  let p = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in`1`
        Bot (n,h,qSelect1 left,qSelect1 right)
  in`2`
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

We finish where we start, splitting \texttt{qSelect} into
\texttt{qSelect1} and \texttt{qSelect2}.  
Omitting a few type annotations, that is:

\begin{lstlisting} 
1`fun qSelect1 (l : list) = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in
        Bot (n,h,qSelect1 left,qSelect1 right)
`2`
fun qSelect2 p (k : int) : int = 
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

The final version, cleaned up and given better datatype names, is shown in \ref{fig:qssplit}.
Observe that \texttt{qSelect1} essentially builds a binary search tree from the input list, 
and \texttt{qSelect2} searches that tree for the correct index.
The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
whereas \texttt{qSelect2} takes $O(\log n)$ time.
Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.

\nr{There's something else I need to say right here before the contribution statement below. 
I want the reader to be thinking "yeah, this splitting business does seem cool."
Bonus points if I also have them thinking "but it would be cooler if it were formal."}

@@

We start by defining a simple staged language, called \lang, to express the input to our splitting algorithm.  
\lang\ unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  
After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
will end up in the stage \bbone\ function ({\tt qSelect1}), 
and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
Typing rules for \lang\ ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.

\begin{figure}

\begin{subfigure}{0.5\textwidth}
\begin{lstlisting}
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...`

datatype tree = Branch of int * int * tree * tree
                | Leaf

1`fun qSelect1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      Branch (n, h, qSelect1 left, qSelect1 right)`

2`fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)`
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\label{fig:qssplit}
\end{subfigure}
\caption{Caption place holder}
\end{figure}

\begin{comment}

%% Umut: this is the earlier version.  I am reorganizing this whole
%% section.

\ur{Indentation seems broken. Be consistent.}

\begin{figure*}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) 
  : (int*list*list) =
  case unroll l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) =>
      let (s,left,right) = partition (p,t) in
      if h<p 
      then (s+1,Cons(h,left),right)
      else (s,left,Cons(h,right))

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\label{fig:quickselect}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
  case l of
    Empty => next {`2`0`1`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        `2`let n = hold{`1`n`2`} in
          case compare prev{`1`k`2`} n of
            LT => prev {`1`qsStaged (left k)`2`}
          | EQ => hold {`1`h`2`}
          | GT => prev {`1`qsStaged (right, 
                             next{prev{k}-n-1)}`2`}`1`
      }`
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\label{fig:qsstaged}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting}
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...`

datatype tree = Branch of int * int * tree * tree
                | Leaf

1`fun qSelect1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      Branch (n, h, qSelect1 left, qSelect1 right)`

2`fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)`
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\label{fig:qssplit}
\end{subfigure}
\caption{Caption place holder}
\end{figure*}

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in \ref{fig:quickselect}.
Quickselect is an algorithm for finding the $k$-th largest element in a list.
It works by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

\ur{I find it hard to distinguish between blue and black on printed paper.}

One notable property of quickselect is that the partitioning of the list, 
even through its recursive calls, depends only on the values of list elements, and not on the value of $k$.
To illustrate this, we present colorized version of \texttt{qSelect}:

\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ... `

fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => `qSelect 1`left `2`k
      | EQ => h
      | GT => `qSelect 1`right `2`(k-n-1)`
\end{lstlisting}

Here, the input $k$ and all parts of the computation that depend on it are colored blue,
whereas all of the parts of the computation that do not depend on $k$ are colored red.
The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.

\ur{I find control dependence issue very confusing.  Let's factor this
out and talk about it in the end. }

To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
first arguments to the recursive calls to \texttt{qSelect} are colored red,
despite appearing underneath a blue case statement.
This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.

Observe the that helper function \texttt{partition} is colored entirely red,
which is valid because it is called with only red arguments.  
This is an illustration of the property stated above, that 
the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.

This suggests an optimization: we can split \texttt{qSelect} into two functions, 
one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
Because of the dependence properties stated above, 
\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
whereas \texttt{qSelect1} will only depend on the input list $l$.
Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
Equationally, we can represent this relationship as simply 
\[
\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
\]
for some list \texttt{l} and index \texttt{k}.

\ur{I don't follow the explanation below, e.g., ``assuming conclusion''}

With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
Because the original function is recursive,
the first step in this process is to assume the conclusion at the call sites.

\ur{quickselect 1 and 2: how did we introduce them?}

This is performed in place:
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}
We then pull the stage \bbone\ portion of each branch of the
comparison up into a single tuple.
\ur{Why did we include h}

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}

\ur{Don't know what ``this last step'' refers to.}
This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
We continue to pull up the precomputation and include more in it:

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

This completes the splitting of the cons branch.  
The empty branch can also be split, though in a trivial way:

\ur{Not clear, ``can be split'' split.  Does or doesn't (split) should be the verb?}
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => let p = () in `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

To continue splitting the case statement, we need to generate a sum type in the precomputation
to remind the code what branch its in.

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  let p = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in`1`
        Bot (n,h,qSelect1 left,qSelect1 right)
  in`2`
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

We finish where we start, splitting \texttt{qSelect} into
\texttt{qSelect1} and \texttt{qSelect2}.  
Omitting a few type annotations, that is:

\begin{lstlisting} 
1`fun qSelect1 (l : list) = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in
        Bot (n,h,qSelect1 left,qSelect1 right)
`2`
fun qSelect2 p (k : int) : int = 
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

>>>>

\ur{
\begin{itemize}
\item 
I find it hard to distinguish between blue and black on printed paper.
\item
$\pause$ should be called ``promote'' or something like that. I
  believe this is what Taha calls cross stage persistence.  THat work should
  be cited.
\item the splitting example should shed more light to the techniques
  and try to give more insight.
\end{itemize}
}

\crem{
Outline:
\begin{enumerate}
\item Say we want to run a bunch of quickselect queries on a single list.
\item We observe that some of the code depends only on the list.
\item We can write this code in \lang\ to make that dependency clear.
\item We can intuitively split into two functions.
\item This is asymptotically better.
\end{enumerate}
}

\input{figures/quickselect}

Staging is a technique for using language constructs and types to
explicitly divide a computation according to the frequency or the
arrival time of data.

Suppose that we wish to issue order statistics queries on a collection of items
represented as a list \texttt{l}. We can use quickselect \cite{quickselect}, an
expected linear time algorithm which takes a list \texttt{l} and a rank
\texttt{k} and returns the element with rank \texttt{k} in \texttt{l}.

Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
by using the first element as a pivot%
\footnote{We assume that the list is prepermuted to guarantee the expected
linear time behavior.}
and then recurs on one of the two resulting sides to find the desired element. The
side chosen is determined by the relationship of \texttt{k} to the size of the
first half \texttt{n}, which is returned by the \texttt{partition} function
along with the two sides.

Now, consider an application where we perform many order statistics queries
on the same collection \texttt{l}, but with $m$ different ranks \texttt{k}:
%
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@.
\end{lstlisting}
%
To take advantage of this, we can \emph{stage} quickselect so that the parts of
the computation that depend on the collection~\texttt{l} is performed once, and
the desired result is computed by using~\texttt{k} and the results from this
prior stage.

%% There has been the extensive research into the question of how to
%% automatically add staging annotations to unstaged code.  This process
%% is known as {\em binding time analysis}, and we do not consider it
%% here.  Instead, we assume that all input programs are properly staged
%% according to some programmer's intent.

\crem{missing text}

Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \ref{fig:qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage.

\crem{explain circle types}

The construct $\next$ occurs in a stage \bbone\ context and indicates that the
contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
\bbtwo\ context and indicates that the contents of its block is stage \bbone.
(We adopt the convention that the top-level context is stage \bbone.)

Informally speaking, the $\next$ construct enables operating on second stage
computations in the first stage essentially by treating them as boxes that can
be moved around, stored in data structures, and passed to functions, and the
$\prev$ construct enables unwrapping the contents of such a box to use in the
second stage.

For convenience, the code also uses $\pause$ blocks.  This construct
has same stage signature as $\prev$, but instead of ``unwrapping''
$\fut$ types it simply promotes integers from stage~\bbone\ to
stage~\bbtwo.%
\footnote{$\pause$ can be implemented from our other language features, with
some effort. We provide it as a primitive to shorten examples. Furthermore, it
would be wise to extend $\pause$ to other base types, if we had them, and to
products and sums thereof.  This is related to the notion of {\em mobility} in
\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

We note that given the type signature of staged quickselect, there are several
ways to insert the staging annotations.  The version we presented here maximizes
the work done at stage~\bbone.

The type system guarantees that all first-stage computations depend only on the
first stage values and can indeed be computed in the absence of second-stage
values.

\crem{partial evaluation requires compile-time access to first stage, whereas we
don't}

\crem{introduce boundary terminology}

Here, we wish to derive two functions, \texttt{quickselect1} and
\texttt{quickselect2}, that are semantically equivalent to quickselect in
the sense that:
\\
\texttt{quickselect l k = quickselect2 (quickselect1 l) k}.
\\
\crem{and moreover they correspond to the stages}

Since \texttt{quickselect1} requires only on a first stage
argument~\texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on~\texttt{l},
e.g.,  the code consisting of $m$ calls 
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@
\end{lstlisting}
%
can be replaced by 
%
\begin{lstlisting}
b = quickselect1 l
quickselect2 b 1
quickselect b 2
@$\ldots$@ 
quickselect b @$m$@.
\end{lstlisting}

@@

\crem{walk through manual example here}

The final version, cleaned up and given better datatype names, is shown in
\ref{fig:qssplit}.  Observe that \texttt{qSelect1} essentially builds a binary
search tree from the input list, and \texttt{qSelect2} searches that tree for
the correct index.

Assuming that the list~\texttt{l} contains $n$ elements, this optimization
changes the asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$
to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.

Note that the desired optimized code shows above is intellectually more
sophisticated than the code that we have started with: the optimized code is
able to create a data structure, a balanced binary tree augmented with indexing
information, and use a binary search technique over this tree to compute the
result asymptotically more efficiently.
%In fact, based our teaching experience, we can imagine this kind of problem to
%be a moderately difficult exam question in an undergraduate algorithms class, as
%it not only requires understanding of data structures such as binary search
%trees but also requires modifying them to augment with indexing information to
%support rank-based search.

@@

\input{figures/quickselect-split}
